---
title: "My First Markdown Document"
author: "Simon Joly and Rutger Vos"
date: "MeBioDA - Fall 2017"
output:
  html_document:
    highlight: haddock
    theme: united
  pdf_document:
    highlight: default
---

R Markdown is a language to easily write dynamic documents from within an R environment. It is based on the very popular and simple Markdown language. It is particularly useful to create documents in which you want to include R code and R figures or results and that are automatically regenerated whenever underlying R code or data changes. R markdown is therefore a useful technology to promote reproducible research. Here we will provide an easy example here on how to create a R Markdown document. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

You will need RStudio installed, as well as the packages `knitr` and `rmarkdown`. You can install them using the command

```
install.packages("rmarkdown")
install.packages("knitr")
```

From RStudio, you can choose `File > New file > RMarkdown` to create a new R Markdown document. The document will have a header that contains a title, the author, the date, and other informations.

```{}
---
title: "My First Markdown Document"
author: "Me"
date: "today"
---
```

You can look on the web for some of the Markdown syntax, but here is a quick example of text, followed by the output:

```
# First header
## Second header
When you write text, you can have it in **bold** or in *italics*.

> This is how you make a block quote
```

# First header
## Second header
When you write text, you can have it in **bold** or in *italics*.

> This is how you make a block quote


However, R Markdown is especially useful for including R code in the output document, as well as the results obtained when the code is executed. For instance, if you want to show the following code in the document:

```{}
summary(cars)
```

It will look like this once formatted with R syntax.

```{r}
summary(cars)
```

As you can see, the command is followed by the R result in the document.

And you can output the results of plot functions such as `plot(cars)` directly in the document:

```{r}
plot(cars)
```

To export your document in html or pdf format, you just press the button "Knit hmtl" or "Knit pdf" in RStudio.

Please try to create a document of your own, reading and plotting a phylogeny for instance.

For more details on using R Markdown see <http://rmarkdown.rstudio.com>. You can also look at the source code for the document presented in this course.
