---
title: "Introduction to phylogenies in R"
author: "Simon Joly and Rutger Vos"
date: "MeBioDA - Fall 2017"
output:
  pdf_document:
    highlight: default
    toc: yes
    toc_depth: 2
  html_document:
    highlight: haddock
    theme: united
    toc: yes
    toc_depth: 2
---

# Introduction

This document introduces working with phylogenies in R. It asssumes that you already know the basics of R. If you don't, you should start by looking at an R tutorial on the web.

This tutorial guides you through different functions to give you an idea of what can be done with phylogenies in R. You will of course learn many more tricks in the following classes.

# R resources

## R Studio

For this class, you need to install R on your computer. You will also need to install [R studio](http://www.rstudio.com/), which is a very practical way to use R. But more importantly, it allows to use KnitR very easily, which you will need for your assignments.

## Packages

There are lots of packages for phylogenetic analyses in R. We won't enumerate them all here, but you will get a good idea of the options available by looking at the [phylogenetic R vignette](http://cran.r-project.org/web/views/Phylogenetics.html) maintained by [Brian O'Meara](http://brianomeara.info/). The vignette is mostly oriented towards phylogenetic comparative methods, but it is a good start.

The most basic package for using trees in R is *ape*, which allows you to read and plot trees, and do some topological analyses.

# Generating, importing, and plotting trees

## Generating a tree

Throughout these exercises, we will often use simulated trees, which are very useful for pedagogical purposes. Trees can be simulated using several functions, but here is an example to simulate one tree with 15 species. (_Question: which key process is missing from this simulator if we want to approximate even the simplest plausible form of diversification?_)

```{r "Simulate Pure-Birth Tree",message=FALSE}
require(phytools)
tree <- pbtree(n=15,nsim=1)
plot(tree)
```

You can save such a tree in [Nexus](http://informatics.nescent.org/w/images/8/8b/NEXUS_Final.pdf) format to a file. But before you do so, it is a good idea to set the working directory to the same folder where your script is saved. You can do that in RStudio in the menu `Session > Set Working Directory > To Source File Location`.

```{r "Save tree file"}
require(ape)
write.nexus(tree, file="My_first_tree.nex")
```

## Simulating characters

Characters can also be easily simulated in R. For instance, you can simulate a continuous character under a [Brownian Motion (BM)](https://en.wikipedia.org/wiki/Brownian_motion) model using the following code:

```{r "Simulate continuous character"}
char1 <- fastBM(tree, sig2=0.01, nsim=1, internal=FALSE)
# To get character values for tree tips:
char1
```

(_Question: if we are looking at form and function, is BM really a realistic model of how functional diversity is generated?_)
Now, let's save this character to a file to pretend it is our original data.

```{r "Write character to file"}
write.table(matrix(char1,ncol=1,dimnames=list(names(char1),"char1")), file="mycharacter.csv", sep=";")
```

Now that we have simulated a tree and a character, let's erase what we have done so far from the R environment and pretend these are our data for the next sections.

```{r "Clear directory"}
rm(list=ls())
```

## Import data into R

Here is how you should import your data into R.

```{r "Import tree and data"}
tree <- read.nexus(file="My_first_tree.nex")
char1 <- read.csv2(file="mycharacter.csv",dec=".")
```

The tree format in ape contains several facets of the data, and it is useful to know how to access them. For instance, the tip labels can be accessed using `tree$tip.label` and the branch lengths using `tree$edge.length`. We will see more options in other exercises, but if you want more detailed information on how objects of type "phylo" are organized, you can have a look at help file `?read.tree` or at this [document](http://ape-package.ird.fr/misc/FormatTreeR_24Oct2012.pdf) prepared by [Emmanuel Paradis](http://ape-package.ird.fr/ep/), the author of `ape`.

## Tree visualization

Plotting trees is one of the very convenient aspects of using R. Options are numerous and possibilities large. The most common function is `plot.phylo` from the ape package, which has a lot of different options. We suggest that you take a close look at the different options of the function `?plot.phylo`. Here is a basic example:

```{r "Basic tree plotting"}
plot(tree, type="c")
add.scale.bar()
```

R is also useful for plotting characters alongside trees. If you have a categorical character, you can use it to color the tips of the phylogeny. In the following example we generate a dummy character by randomly sampling states from a three-state character. (_Question: do you think this character will "fit" the tree well? Why (not)?_)

```{r "Plot categorical character"}
# Generate a random categorical character
char2 <- as.factor(sample(c(1,2,3),size=length(tree$tip.label),replace=TRUE))
# Create color palette
library(RColorBrewer)
ColorPalette1 <- brewer.pal(n = length(levels(char2)), name = "Dark2")
plot(tree, type="p", use.edge.length = TRUE, label.offset=0.2,cex=1)
tiplabels(pch=21,bg=ColorPalette1[char2],col="black",cex=1,adj=0.6)
op<-par(xpd=TRUE)
legend(0,0,legend=levels(char2),col=ColorPalette1,
       pch=20,bty="n",cex=1,pt.cex=1.5,ncol=length(levels(char2)))
par(op) #reset graphical parameters to defaults
```

A similar result can be obtained with a continuous variable. Here, we will use the Brownian Motion data: 

```{r "Plot continuous character"}
# Discretize continuous character into 8 categories
char1.cat <- cut(char1[,1],breaks=8,labels=FALSE)
# Create color palette
ColorPalette2 <- brewer.pal(n = 8, name = "Blues")
# Plot the tree
plot(tree, type="p", use.edge.length = TRUE, label.offset=0.2,cex=1)
tiplabels(pch=21,bg=ColorPalette2[char1.cat],col="black",cex=1,adj=0.6)
op<-par(xpd=TRUE)
legend(0,0,legend=levels(cut(char1[,1],breaks=8)),
       col=ColorPalette2,pch=20,bty="n",cex=0.7,pt.cex=1.5,ncol=4)
par(op) #reset graphical parameters to defaults
```

As expected from a character simulated with Brownian motion, you can see that closely related species tend to have more similar character values. (_Question: was that also the case for the three-state character?_)

Another way to plot a continuous character is to use the function `table.phylo4d` from the `adephylo` package to visualize the character such that its values are represented by circles of different sizes and colors. With this package it is also possible to plot multiple characters at the same time, as shown here by simulating another BM character. (_Question: which of the two BM characters (char1 or char3) will have higher overall variance?_)

```{r "Plot continuous character - table.phylo4d", message=FALSE}
library(phylobase)
library(adephylo)
char3 <- fastBM(tree, sig2=0.1, nsim=1, internal=FALSE) # simulate a faster evolving character
char.table <- data.frame(char1=char1[,1], char3)
obj <- phylo4d(tree, char.table) # build a phylo4d object
op <- par(mar=c(1,1,1,1))
table.phylo4d(obj,cex.label=1,cex.symbol=1,ratio.tree=0.8,grid=FALSE,box=FALSE)
par(op) #reset graphical parameters to defaults
```

You can also visualize the character as a "phenogram". (At least, that is what the `phytools` calls this visualization. This is not necessarily a term that is universally agreed upon - but what it shows is the character values along the Y-axis of the tree):

```{r "phenogram"}
require(phytools)
phenogram(tree,char3,spread.labels=TRUE)
```

Finally, it is also possible to represent a tree on a 2-dimensional plot, coloring points with the categorical variable.

```{r "Phylomorphospace", message=FALSE}
phylomorphospace(tree,char.table)
points(char.table,pch=21,bg=ColorPalette1[trait2],col="black",cex=1.2,adj=1)
legend("topleft",legend=levels(char2),
       col=ColorPalette1,pch=20,bty="n",cex=1,pt.cex=1.5,ncol=length(levels(char2)))
```

# Handling multiple trees

Under numerous circumstances it will be important to know how to handle multiple trees in R. These are normally stored in a `multiPhylo` object. Let's see an example.

```{r "Simulate several trees"}
trees <- pbtree(n=15,nsim=10)
trees
```

You can see that the object is not the same as a phylo object. For instance, if you use the code `plot(trees)`, you will be prompted to hit enter to page from one tree to the other. To plot the individual trees, you need to use the following technique:

```{r "multiPhylo manipulation"}
trees[[1]]
plot(trees[[1]])
```

# Manipulating trees

There are several manipulations that can be made to trees. Here are a few examples.

## Drop tips

```{r "Drop tips"}
plot(drop.tip(tree,c("t1","t2","t3","t4","t5")))
```

## Reroot trees

```{r "Reroot trees"}
plot(root(tree,"t1"))
```

## Get cophenetic distances

```{r "Cophenetic distance"}
cophenetic.phylo(tree)
```

# Ploting posterior probability values

The package `phyloch` has very useful functions to read trees from BEAST or MrBayes. Inconveniently, it has to be installed from the binaries available from [Christoph Heibl's webpage](http://www.christophheibl.de/Rpackages.html). Once you have downloaded the package, you can install it using RStudio from the menu 'Tools > Install Packages' and select the package that you have just downloaded. You also need to install the package `XML`, which you can do using the command `install.packages("XML")`.

## Read Bayesian trees

The functions `read.beast` and `read.MRBayes` parse chronograms in NEXUS format as produced by TreeAnnotator or output by MrBayes, respectively. Here is an example with a tree obtained from TreeAnnotator.

```{r "Read BEAST tree", message=FALSE, echo=FALSE, fig.height=11, fig.width=9}
library(phyloch)
beasttree <- read.beast("./data/seedplants.tre")
plot(beasttree)
node.support(beasttree$posterior,cutoff=0,digits=2)
```

> If you want to drop tips from trees in phyloch packages, use the function `drop.tip2` from the `phyloch` package, and not the `drop.tip` from the `ape` package.

The function `read.beast` give a lot of annotations to the tree, which you can see using the command `str(tree)`. For instance, you can access the posterior probability values using `beasttree$posterior`.

## Plot node bars

The `phyloch` package can also plot bars representing the highest posterior density (HPD) intervals of node ages.

```{r "Plot HPD of node ages", fig.height=11, fig.width=9}
plot(beasttree)
HPDbars(beasttree, label = "height_95%_HPD")
```


Info on Rmarkdown: <http://rmarkdown.rstudio.com>.
